---
title: "Alpha-taxonomy in world mammals"
#subtitle: "Trends in methods, species concepts, and speciation study from the 1970s to today"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
fontsize: 12pt
geometry: margin = 0.75in, a4paper
mainfont: Times New Roman
csl: harvard-university-of-cape-town-rvm_2017-08-15.csl
bibliography: Spp_speciation_deliverable.bib
header-includes:
  - \usepackage{lipsum}
  # Institutional logos as a wrapping figure
  - \usepackage{wrapfig}
  # Hanging indentation for references §
  - \usepackage{hanging}
  # Formatting
  - \setlength{\parindent}{30pt}
  - \setlength{\parskip}{0pt}
  - \renewcommand{\baselinestretch}{1.5}
---

```{r msw3-script-setup, include = FALSE}
rm(list = ls())
if (!require(pacman)) install.packages("pacman")
library(pacman)
p_load(here)
here()
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
knitr::opts_knit$set(root.dir = paste(here()))
```

```{r source-msw3-script}
# FIXME
source("msw3_analyses.R")
p_load(tidyverse, magrittr, here, stringr, glue, rjson)
```


\begin{wrapfigure}{r}{100pt}  % this figure will be at the right
    \includegraphics[width=1.5cm, height=1.5cm]{uct_logo.png} \hfill
    \includegraphics[width=1.5cm, height=1.5cm]{biosci_logo_crop.png}
\end{wrapfigure}

\par \noindent \textbf{R. van Mazijk} VMZRUA001@myuct.ac.za

\footnotesize
\par \noindent `r format(Sys.time(), "%Y-%m-%d")`
\par \noindent BSc Hons Biological Sciences, Species & speciation module deliverable
\par \noindent For A/Prof A.M. Muasya

\par \noindent \hrulefill

\hfill


\normalsize

\noindent Herein I discuss the progress of mammalian alpha-taxonomy from 1975 to 2005<!--and up to 2017-->. First, I focus on the methods employed to diagnose and delimit these mammal species and how these have changes over the the past 30 to 40 years. Second, I discuss the new species and groups discovered in this period. Third, I will attempt to draw conclusions, if possible, about the speciation mechanisms at work in mammal populations based on the evidence use to delimit those mammal species. I will frame these discussions in the context of the species concepts (or lack thereof) typically used in mammalogy and mammalian taxonomy.


## Mammalogical species concepts

The species concept or criteria a taxonomist follows can affect the number of species delimited. Supposing there is reality to species, and that two lineages being different species is grounded in reality, species concepts enable us to estimate when this occurs. Following the reasoning of @DeQueiroz2011, different species concepts each deal with different lines of empirical evidence. These each describe different aspects of the speciation process, such that in each the criteria for delimiting two different species is based on some property of each lineage that would differ between them if they were separate species [@DeQueiroz2011]. Consequently, different species concepts each encapsulate *some* aspect of dissimilarity between two lineages as a consequence of speciation. @DeQueiroz2011 argues that should any one criterion be met, it is parsimonious to assume speciation has begun, and that the bifurcation of the lineages be appropriately regarded with a species delineation. This aside, consistency in species concepts in taxonomic research is still desirable (if perhaps unattainable<!--ref from module readings-->). The case with mammal taxonomists is summarised well:

\small

> \textit{Mammalogists, for the most part, have avoided debate on species concepts, but most investigators follow a particular species definition.}

\normalsize

\noindent @Baker2006 make this somewhat alarming statement about the *modus operandi* of mammal taxonomists. It appears that mammal taxonomy, like many fields of biology, has agreed-upon nomenclatural standards, but has little consensus on which species concepts to use when delineating species. This precludes ready comparison of, say, mice species vs bovid species, as there is no guarantee that "species" represent similarly biologically important units. <!--Perhaps this isn't a bad thing? But I suppose that depends on the taxonomic scale at which SCs fail at facilitating comparison-->

<!--elaborate: Idiosyncratic/researcher-specific species concept usage...-->

For the most part, mammologists historically<!--when?--> followed (if only loosely) the Biological Speciec Concept (BSC)<!--ref Mayr-->. This is intuitive as mammals are sexually reproducing eukaryotes [@Mayr; @Baker2006]<!--elaborate-->. The Phylogenetic Species Concept (PSC) contrasts with the BSC in focussing on an explicitly evolutionary, lineage based criterion<!--ref--> <!--elaborate on allopatry solution that the PSC gives-->. The PSC thus now dominates in mammalogy, as the most common species concept used or referred to<!--ref-->. The PSC solves the problem of the BSC wherein entities widely regarded as ideally being two species fail to be differentiated reproductively. @Baker2006 themseleves put forward the Genetic Species Concept (GSC), wherein a species represents:

\small

> \textit{[...] a group of genetically compatible interbreeding natural populations that is genetically isolated from other such groups.}

\normalsize

\noindent This does not necessarily imply reproductive isolation either, as can be seen in the case of *Thomomys* pocket gophers [@Baker2006]<!--elaborate-->. The transition from majority usage of the BSC to PSC in mammal taxonomy is not so simple, however. There is little explicit consideration of species concepts by the majority of mammal taxonomists<!--ref-->. And, there is debate. @Groves2012 in particular advocate more strongly for the PSC, even in place of the GSC, in that species defined by the PSC are:

\small

> \textit{[...] 100\% diagnosable: they have fixed heritable differences between them [...] [and the PSC] offer[s] a repeatable, falsifiable definition of species [...]}

\normalsize

\noindent This is evident in the widespread application of the PSC in mammal taxonomy. But nonetheless, the lack of explicit concern for species criteria is troubling<!--elaborate-->. Do mammal taxonomists employ similar methodogies when delimiting species? I discuss this below, despite the nuanced inconsistencies that differing species concepts among mammalogists may lead to.


## Lines of evidence

Morphology was the dominant method in mammal species delimitation prior to 1985 [@Baker2006]. Indeed, morpholgical differences between taxa are still influential in their taxonomy. Furthermore, allopatric populations of mammal that are morphologically distinct (whatever that may entail) are regularly treated as separate species. This concurs with the usage of the PSC above, where non-reproductively isolated taxa (e.g. grizzly and polar bears; lions and tigers) are treated as separate species<!--elaborate-->. Phenetics still common practice in mammal taxonomy, wherein morphological divergence between populations is numerically described. Species are then diagnosed as clusters in this quantitative trait-space<!--more!-->. Phenetics and morphology are often also the starting points of alpha-taxonomic research [e.g. @Lew2006; @Bronner2000], as it is more amenable to efficient, cheap study than molecular systematics.

Early molecular work used in mammal taxonomy employed karyotype studies [e.g. @Bronner2000]---to delineate species on the basis of chromosomal traits [@Baker2006] (e.g. diploid number, chromosome structure, arrangement). Allozymes and DNA sequencing methods arrived later. These molecular methods were costly in their infancy [@Baker2006], but soon became much more practical. The benefits of these methods is their greater objectivity, in that qualitatvely scored morphological traits are more subject to researchers' idiosyncrasies<!--ref--><!--doesn't apply as much to phenetics-->. The other benefit of characters divorced from observable morphology is that they can allow cryptic species to be uncovered [@Zimmermann2014]. This does, however, contribute to taxonomic inflation.<!--ref; is this bad? elaborate--> As @Rylands2012 point out (in reference to primates, but this easily pertains to many biota), taxonomic inflation may indeed represent a greater understanding of existing diversity<!--elaborate-->. Furthermore, molecular systematic methods may reveal morphologically distinct populations to in fact be polymorphs of a single species [@Zimmermann2014]. In fact, 


## Mammal discoveries, 1975---2005

*Mammal Species of the World, 3rd Edition* (MSW3) [@MSW3] is a reputable<!--ref?--> database of mammalian taxa (species, sub-specific, and super-specific) and information concerning their treatments (e.g. scientific name, authors' name and year described, original publication citation, distribution, etc. [@MSW3]) from 1702 to 2005 AD. This data product is a boon to the discussions here. Using the `.csv` file from the MSW3 webpage (<https://www.departments.bucknell.edu/biology/resources/msw3/>), I explored this dataset in `R` [@RCoreTeam2017], primarily using the `tidyverse` suite of packages for data exploration and visualisation [@Wickham2017]. Not only does MSW3 enable exploration of mammal taxonomic discovery through time, it also facilitated easier literature searches, providing common journals in which mammal descriptions are published.

There is a subset of mammal order that are most taxonomically "active" (Figure \ref{taxa-thru-time}), such that there are only eight orders in which $\geq 10$ new species were described in the period 1975---2005. This is worth bearing in mind when discussing mammalian taxonomy; there is a taxonomic bias to, either, the rate of discovery in certain taxa, or the species richness of certain taxa. Either way, species concepts have bearing on these numbers, discussed below.

![\label{taxa-thru-time} \footnotesize Bar-plots of the number of new species (\textit{sensu stricto}, excluding sub-specific additions) described each year in the eight most taxonomically active orders of mammal over the period 1975---2005. Data are from MSW3 (see text) [@MSW3]. Taxonomically "active" here implies that $\geq 10$ new species were described for an order over the 1975---2005 period. These eight orders are those that meet this criterion, such 533 species' discoveries are illustrated here. For reference, Dasyuromorphs are commonly referred to as Tasmanian devils and numbats, Didelphimorphs as opossums, Diprotodonts as koalas and wombats, and Soricomorphs as moles and shrews](ppt_fig.png){ width=65% }

![\label{taxa-thru-africa} \footnotesize Bar-plots of the number of new species (\textit{sensu stricto}, excluding sub-specific additions) described each year in the three orders of mammal over the period 1975---2005 that had *any* new species found in that period. Not shown is Afrosoricidimorpha, which had one new species: *`r paste(SA_sl_msw$Genus, SA_sl_msw$Species)[1]`*, a golden mole [@Bronner2000] (see text). Data are from MSW3 (see text) [@MSW3].](ppt_fig_africa.png){ width=65% }

Discoveries of new mammal species over this 30 year period are dominated by rodents and bats (Figure \ref{taxa-thru-time}). It is thus interesting that, in the Cape, only one new mammal species was described in the the period 1975 to 2005 [@Meester1978]: *`r paste(cape_sl_msw$Genus, cape_sl_msw$Species)`*, from the order `r cape_sl_msw$Order`, discovered around Knysna. This shrew species was first delimited on the basis of its morphology and geographic range, distinct from those of its relatives [@Meester1978].

South African mammal species over this period include *`r paste(cape_sl_msw$Genus, cape_sl_msw$Species)`*, the rodent *`r paste(SA_sl_msw$Genus, SA_sl_msw$Species)[2]`*, and the golden mole *`r paste(SA_sl_msw$Genus, SA_sl_msw$Species)[1]`* (Afrosoricidimorpha) [@Bronner2000]. In the case of the golden mole, karyotypic and phenetic methods were employed, as mentioned above. African mammal fauna discoveries are more diverse: 23 new species in the period 1975---2005, including primates, rodents, shrews, and indeed *`r paste(SA_sl_msw$Genus, SA_sl_msw$Species)[1]`* (Figure \ref{taxa-thru-africa}). It is intuitive that Africa as a whole would have yielded primate discoveries, compared to only South Africa, as there are fewer primate species outside of the tropics. Globally, 


- Africa
    - Primates, Rodents, Soricom.
- Global?
    - <more>
- e.g. cytochrome-b (mtDNA)
    - Carvalho et al. 2009. Phylogeny of Thylamys (Didelphimorphia, Didelphidae) species, with special reference to Thylamys karimii
- cytochrome-b in rodents
    - Re: GSC sensu Baker & Bradley 2006 (J. of Mammal. 87(4):643–662), Bradley & Baker 2001 (J. of Mammal. 82(4):960–973)
- multi-locus work in Afro-tropical shrews
    - cytochrome-b, among others,
    - Jacquet et al. 2012. Zool. J. Linn. Soc. 166(3):672–687


## Modes of mammalian speciation



## Concluding remarks

- Molecular species delimitation is relevant to PSC/ESC/BSC
- Ecologically relevant speciation processes (e.g. bats) are case-by-case
- Little generality to be drawn about speciation from the alpha-taxonomic methods 
- Genetic sequencing & PSC (or at least ESC)
- Good taxonomic inflation!
- Genomics is up-and-coming! e.g. Murphy et al. 2004


\noindent \hrulefill

# \small References
\footnotesize \hangparas{15pt}{1}
