---
title: "Alpha-taxonomy in world mammals"
#subtitle: "Trends in methods, species concepts, and speciation study from the 1970s to today"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
fontsize: 12pt
geometry: margin = 0.75in, a4paper
mainfont: Times New Roman
csl: harvard-university-of-cape-town-rvm_2017-08-15.csl
bibliography: Spp_speciation_deliverable.bib
header-includes:
  - \usepackage{lipsum}
  # Institutional logos as a wrapping figure
  - \usepackage{wrapfig}
  # Hanging indentation for references §
  - \usepackage{hanging}
  # Formatting
  - \setlength{\parindent}{30pt}
  - \setlength{\parskip}{0pt}
  - \renewcommand{\baselinestretch}{1.5}
---

```{r msw3-script-setup, include = FALSE}
rm(list = ls())
if (!require(pacman)) install.packages("pacman")
library(pacman)
p_load(here)
here()
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = here())
```

<!--Can't cache, something in `magrittr::not()` breaks?-->
```{r source-msw3-script, include = FALSE, cache = FALSE}
source("msw3_analyses.R")
p_load(tidyverse, magrittr, here, stringr, glue, rjson)
```


\begin{wrapfigure}{r}{50pt}  % this figure will be at the left
    \par \includegraphics[width=1.25cm, height=1.25cm]{uct_logo.png}
    \par \includegraphics[width=1.25cm, height=1.25cm]{biosci_logo_crop.png}
\end{wrapfigure}

\par \textbf{R. van Mazijk}
\par VMZRUA001@myuct.ac.za

\footnotesize
\par \noindent `r format(Sys.time(), "%Y-%m-%d")`
\par \noindent BSc Hons Biological Sciences
\par \noindent Species & speciation module deliverable
\par \noindent For A/Prof A.M. Muasya

\par \noindent \hrulefill


\normalsize

## Introduction

Herein I discuss the progress of mammalian alpha-taxonomy from 1975 to 2005<!--and up to 2017-->. First, I discuss the new species and groups discovered in this period. Second, I focus on the methods employed to diagnose and delimit these mammal species and how these have changes over the the past 30 to 40 years. Third, I will attempt to draw conclusions, if possible, about the speciation mechanisms at work in mammal populations based on the evidence use to delimit those mammal species. And lastly, I will frame these discussions in the context of the species concepts (or lack thereof) typically used in mammalogy and mammalian taxonomy.

*Mammal Species of the World, 3rd Edition* (MSW3) [@MSW3] is a reputable<!--ref?!--> database of mammalian taxa (species, sub-specific, and super-specific) and information concerning their treatments (e.g. scientific name, authors' name and year described, original publication citation, distribution, etc. [@MSW3]) from 1702 to 2005 AD. This data product is a boon to the discussions here. Using the `.csv` file from the MSW3 webpage (<https://www.departments.bucknell.edu/biology/resources/msw3/>), I explored this dataset in `R` [@RCoreTeam2017], primarily using the `tidyverse` suite of packages for exploration and visualisation [@Wickham2017].

There is a subset of mammal order that are most taxonomically "active" (Figure \ref{taxa-thru-time}), such that there are only eight orders in which $≥10$ new species were described in the period 1975---2005.

```{r taxa-thru-time, fig.height = 4, fig.width = 4, fig.cap ='\\label{taxa-thru-time} \\footnotesize  (ca. 54 hours range of sampling): \\normalsize \\textbf{A)} \\footnotesize Rain-fall input ($m^{3} h^{-1}$), \\normalsize \\textbf{B)} \\footnotesize stream-flow rate ($m^{3} h^{-1}$) of the Liesbeeck River, and \\normalsize \\textbf{C)} \\footnotesize $\\delta^{18}O$ (\u2030) and \\normalsize \\textbf{D)} \\footnotesize $\\delta^{2}H$ (\u2030) values for both the rain- and stream-water (keyed).'}
# TODO: add ordinal silhouettes to facets?
plot_new_taxa(msw, "SPECIES",
    by = "Order",
    which_orders = get_the_coolest(msw, "Order"),
    ylab = "spp.",
    drop_fill = TRUE,
    drop_facet = TRUE)
```

## Mammal discoveries, 1975---2005

## Lines of evidence

## Modes of mammalian speciation

## Mammalogical species concepts

<!--

<fig. 1 focal orders>

“Mammalogists, for the most part, have avoided debate on species concepts, but most investigators follow a particular species definition”; —Baker & Bradley 2006 (J. Mammal. 87(4):643–662)

BSC -> PSC } entirely not this simple; Little explicit consideration from majority of mammalogists

-> GSC (Genetic Species Concept Baker & Bradley 2006 (J. Mammal. 87(4):643–662)); “a group of genetically compatible interbreeding natural populations that is genetically isolated from other such groups”; ≠ reproductive isolation (necessarily); e.g. Thomomys pocket gophers; But there is debate... “[…] 100% diagnosable: they have fixed heritable differences between them […]”; “[…] offer a repeatable, falsifiable definition of species […]”; Groves 2012. American Journal of Primatology. 74(8):687–691.

Lines of evidence in mammal species delimitation; 1) Morphology (pre-1985) (But still influential! e.g. in allopatry); Phenetics still common, or 1st port-of-call e.g. Lew et al. 2006. J. Mammal. 87(2):224–237; 2-4) Karyotype studies, Allozymes, DNA sequencing; costly in the early days, but soon very doable!! very objective!; can highlight cryptic species; taxonomic infaltion?

Cape
Myosorex longicaudatus
	Meester & Dippenaar 1978. Annals of the Transvaal Museum. 31(4):30–42
	Knysna
	Morphology & geography

Africa
Primates, Rodents, Soricom.

Global?
<more>
e.g. cytochrome-b (mtDNA) Carvalho et al. 2009. Phylogeny of Thylamys (Didelphimorphia, Didelphidae) species, with special reference to Thylamys karimii

“The increase in the number of named species […] [means] that we have increased our perception and understanding of the diversity of the primates.” Rylands et al. 2012. International Zoo Yearbook. 46(1):11–24

cytochrome-b in rodents
Re: GSC sensu Baker & Bradley 2006 (J. of Mammal. 87(4):643–662),
	        Bradley & Baker 2001 (J. of Mammal. 82(4):960–973)

multi-locus work in Afro-tropical shrews
cytochrome-b, among others,
	Jacquet et al. 2012. Zool. J. Linn. Soc. 166(3):672–687

Conclude:
Molecular species delimitation is relevant to PSC/ESC/BSC
Ecologically relevant speciation processes (e.g. bats) are case-by-case
Little generality to be drawn about speciation from the alpha-taxonomic methods 
Genetic sequencing & PSC (or at least ESC)
Good taxonomic inflation!
Genomics is up-and-coming! e.g. Murphy et al. 2004

-->

## Concluding remarks


\noindent \hrulefill

<!--Figures-------------------------------------------------------------------->

e.g.
    `Blah blah blah (Figures \ref{o_vs_h}B, \ref{o_vs_h}C).`
    
e.g.
`{r timeline, warning = F, fig.height = 7, fig.width = 7, fig.cap ='\\label{timeline} \\footnotesize Timeline of isotopic and hydrological values throughout the winter storm-event during which sampling was carried out (ca. 54 hours range of sampling): \\normalsize \\textbf{A)} \\footnotesize Rain-fall input ($m^{3} h^{-1}$), \\normalsize \\textbf{B)} \\footnotesize stream-flow rate ($m^{3} h^{-1}$) of the Liesbeeck River, and \\normalsize \\textbf{C)} \\footnotesize $\\delta^{18}O$ (\u2030) and \\normalsize \\textbf{D)} \\footnotesize $\\delta^{2}H$ (\u2030) values for both the rain- and stream-water (keyed).'}`

<!--/Figures------------------------------------------------------------------->


\noindent \hrulefill

# \small References
\footnotesize \hangparas{15pt}{1}
